---
title: "ISLR Chapter 7"
output: html_document
# https://www.youtube.com/watch?v=uQBnDGu6TYU
---

Splines
-------
Splines are more flexible than polynomials, but the idea is rather similar.
Here we will explore cubic splines. 

```{r}
require(ISLR)
attach(Wage)
require(splines)
fit=lm(wage~bs(age, knots=c(25,40, 60)), data=Wage)
plot(age,wage,col='darkgrey')
lines(age.grid,predict(fit,list(age=age.grid)), col='darkgreen',lwd=2)
abline(v=c(25,40,60), lty=2,col='darkgreen')
```

The smoothing splines does not require knot selection, but it does have a smoothing parameter, which can conveniently be specific via the effective degrees of freedom or 'df'.

```{r}
fit=smooth.spline(age,wage,df=16)  #degress of freedom
lines(fit,col='red',lwd=2)
```

Or we can use LOO cross-validation to select the smoothing parameter for us automatically:

```{r}
fit=smooth.spline(age,wage,cv=TRUE)
lines(fit,col='purple',lwd=2)
fit
```
Generalized Additive Models
---------------------------

So far we have focused on fitting models with modely single nonlinear terms. The 'gam' package makes it easier to work with multilpe nonlinear terms. In addition it knows how to plot these functions and their standard errors. 

```{r fig.width=10, fig.height=5}
#install.packages('gam')
require(gam)
gam1=gam(wage~s(age,df=4)+s(year,df=4)+education, data=Wage)
par(mfrow=c(1,3))
plot(gam1, se=T)
gam2=gam(I(wage>250)~s(age,df=4)+s(year,df=4)+education,data=Wage,family=binomial)
plot(gam2)
```

Let's see if we need a nonlinear term for year

```{r}
gam2a=gam(I(wage>250)~s(age,df=4)+year+education,data=Wage,family=binomial)
anova(gam2a,gam2,test='Chisq')
```
One nice feature of the 'gam' package is that it knows how to plot the functions nicely, even for models fit by 'lm' and 'glm'.

```{r fig.width=10, fig.height=5}
par(mfrow=c(1,3))
lm1=lm(wage~ns(age,df=4)+ns(year,df=4)+education,data=Wage)  #ns=natural spline
plot.gam(lm1, se=T) # use plot method gam
```


